### 2020-06-05 writeup/intro.tex

Our main result is to develop tight bounds for the bias and variance of the multi-task estimator for two tasks.
Since the total sample size has increased, the variance of the multi-task estimator is \textit{always smaller} than that of the single-task estimator.
Since task models are different, the bias of the multi-task estimator is \textit{always larger} than that of the single-task estimator (which is $0$ since $n_1 > p$).
Hence, the tradeoff between bias and variance determines whether the transfer is positive or negative.
In Lemma \ref{lem_cov_shift_informal}, we provide a sharp analysis of the variance that scales only with sample size and covariate shift and extend it to the bias.
Combined together, we analyze the bias-variance tradeoff for two tasks in Theorem \ref{thm_main_informal} and extend the analysis to many tasks with same features in Theorem \ref{thm_many_tasks}.
%For the setting of two tasks, we show how the variance of the multi-task estimator  scales with sample size and covariate shift in the following result.
%\textit{Our first contribution} is to develop a concentration bound that arises naturally from the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for two tasks.
%Let $\hat{\beta}_t^{\STL}$ denote the single-task estimator.
%Without loss of generality, let the $t$-th task denote the target task.
%Importantly, the target task's data size is a fixed constant times $p$ in the high-dimensional setting.
%Hence adding more labeled data can help improve its test performance.
%$B\in\real^{p\times r}$
%$\set{W_i \in \real^{r}}_{i=1}^t$

@@

Next, we explain the performance of multi-task learning by varying the three components above in isotropic and covariate shifted settings.
We describe three simulations in Figure \ref{fig_model_shift_phasetrans} for two linear regression tasks and measure the loss of the single-task estimator minus the multi-task estimator.
%We observe that negative transfer occurs as (a) \textit{task similarity}: tasks become more different; (b) \textit{data size}: source/target data size increases.
%\textbf{Task similarity:}
%\textbf{Data sizes:}
%\textbf{Covariate shift:}
%Furthermore, MTL performance is negatively affected when (c) \textit{covariate shift}: the covariance matrices of the two tasks become more different.
\squishlist
	\item In Proposition \ref{prop_dist_transition}, we provide a deterministic condition that depends on the sample sizes and model distance in the isotropic setting.
	Our result predicts most of the empirical observations in Figure \ref{fig_model_shift} correctly.
%	It is crucial that the concentration result in Theorem \ref{lem_cov_shift_informal} is sufficiently precise so that we can explain the transition phenomena in Figure \ref{fig_model_shift} and \ref{fig_size}.
	The unexplained observations are caused by an error term that arises from the bias of the multi-task estimator.
	We discuss these in Section \ref{sec_insight} more precisely.
	\item In Proposition \ref{prop_data_size}, we provide a similar condition that explains a curious phenomenon in Figure \ref{fig_size}, namely adding more labeled data from the source task does not always improve performance.
	%Theorem \ref{lem_cov_shift_informal} allows us to compare MTL performance under different covariate shifts.
	\item In Proposition \ref{prop_covariate}, we study a special case of having the same task models and show that MTL is best performing without covariate shift.
	For this special case, the bound of Lemma \ref{lem_cov_shift_informal} matches the empirical curve in Figure \ref{fig_covariate} accurately.
\squishend
%In Section \ref{sec_insight}, we consider three components including task similarity, data size and covariate shift for a simplified isotropic setting of two tasks.
%We measure task similarity by how small is the distance between $\beta_1$ and $\beta_2$.
%Using our tool, we explain a transition from positive to negative transfer as task similarity decreases.
%		Furthermore, we show that negative transfer is more likely to occur when the source task labels are particularly noisy.
%		In Section \ref{sec_validate}, we validate the observation on text and image classification tasks.
%	In , we provide the trade-off between $\norm{\beta_1 - \beta_2}^2$ and a certain function $\Phi(\rho_1, \rho_2)$ to determine the type of transfer.
%We show that increasing the data size of the source task does not always improve performance for the target task in multi-task learning.
Along the way, we explain a key empirical finding of Taskonomy by Zamir et al. \cite{ZSSGM18}, by analyzing the benefit of MTL for reducing labeled data to achieve comparable performance to STL, .
%We show that covariate shift, measured by $\Sigma_1^{1/2}\Sigma_2^{-1/2}$, is another cause for suboptimal performance for $\hat{\beta}_t^{\MTL}$.
%		We show that as $n_1 / n_2$ becomes large, having no covariate shift between the source and target tasks yields the optimal performance for the target task.
%		On the other hand, when $n_1 / n_2$ is small, there are counter examples where having the same covariance matrix is not necessarily the optimal choice.

Finally, we connect our theory to practical problems of interest.
We provide a single-task based metric to predict positive or negative transfer in multi-task learning.
While it is not well understood when multi-task learning provides positive transfer, we show that the STL results can help indicate and understand MTL results on ChestX-ray14 \cite{chexnet17} and sentiment analysis datasets \cite{LZWDA18}.
Inspired by our observation that adding more source task data does not always improve perfromance, we propose to train multi-task models by incrementally adding source task data for predicting a target task.
We show that our training schedule reduces the computational cost by $55\%$ compared to baseline multi-task training on the sentiment analysis dataset, while keeping the accuracy the same.
Last, we validate our three theoretical findings on sentiment analysis tasks.

>>>>

\textbf{Main results.} First, we develop tight bounds for the bias and variance of the multi-task estimator for two tasks.
Since the total sample size has increased, the variance of the multi-task estimator is \textit{always smaller} than that of the single-task estimator.
Since task models are different, the bias of the multi-task estimator is \textit{always larger} than that of the single-task estimator (which is $0$ since $n_1 > p$).
Hence, the tradeoff between bias and variance determines whether the transfer is positive or negative.
We provide a sharp analysis of the variance that scales only with sample size and covariate shift in Lemma \ref{lem_cov_shift_informal} and extend it to the bias.
Combined together, we analyze the bias-variance tradeoff for two tasks in Theorem \ref{thm_main_informal} and extend the analysis to many tasks with same features in Theorem \ref{thm_many_tasks}.
%For the setting of two tasks, we show how the variance of the multi-task estimator  scales with sample size and covariate shift in the following result.
%\textit{Our first contribution} is to develop a concentration bound that arises naturally from the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for two tasks.
%Let $\hat{\beta}_t^{\STL}$ denote the single-task estimator.
%Without loss of generality, let the $t$-th task denote the target task.
%Importantly, the target task's data size is a fixed constant times $p$ in the high-dimensional setting.
%Hence adding more labeled data can help improve its test performance.
%$B\in\real^{p\times r}$
%$\set{W_i \in \real^{r}}_{i=1}^t$

@@

Second, we explain the performance of multi-task learning by varying the three components above in isotropic and covariate shifted settings.
We describe three simulations in Figure \ref{fig_model_shift_phasetrans} for two linear regression tasks and measure the loss of the single-task estimator minus the multi-task estimator.
%We observe that negative transfer occurs as (a) \textit{task similarity}: tasks become more different; (b) \textit{data size}: source/target data size increases.
%\textbf{Task similarity:}
%\textbf{Data sizes:}
%\textbf{Covariate shift:}
%Furthermore, MTL performance is negatively affected when (c) \textit{covariate shift}: the covariance matrices of the two tasks become more different.
\squishlist
	\item We provide a deterministic condition that depends on the sample sizes and model distance in the isotropic setting (Proposition \ref{prop_dist_transition}).
	Our result predicts most of the empirical observations in Figure \ref{fig_model_shift} correctly.
%	It is crucial that the concentration result in Theorem \ref{lem_cov_shift_informal} is sufficiently precise so that we can explain the transition phenomena in Figure \ref{fig_model_shift} and \ref{fig_size}.
	The unexplained observations are caused by an error term that arises from the bias of the multi-task estimator.
	We discuss these in Section \ref{sec_insight} more precisely.
	\item We provide a similar condition that explains a curious phenomenon in Figure \ref{fig_size}, namely adding more labeled data from the source task does not always improve performance (Proposition \ref{prop_data_size}).
	%Theorem \ref{lem_cov_shift_informal} allows us to compare MTL performance under different covariate shifts.
	\item We study a special case of having the same task models and show that MTL is best performing without covariate shift (Proposition \ref{prop_covariate}).
	For this special case, the bound of Lemma \ref{lem_cov_shift_informal} matches the empirical curve in Figure \ref{fig_covariate} accurately.
\squishend
%In Section \ref{sec_insight}, we consider three components including task similarity, data size and covariate shift for a simplified isotropic setting of two tasks.
%We measure task similarity by how small is the distance between $\beta_1$ and $\beta_2$.
%Using our tool, we explain a transition from positive to negative transfer as task similarity decreases.
%		Furthermore, we show that negative transfer is more likely to occur when the source task labels are particularly noisy.
%		In Section \ref{sec_validate}, we validate the observation on text and image classification tasks.
%	In , we provide the trade-off between $\norm{\beta_1 - \beta_2}^2$ and a certain function $\Phi(\rho_1, \rho_2)$ to determine the type of transfer.
%We show that increasing the data size of the source task does not always improve performance for the target task in multi-task learning.
Along the way, we explain a key empirical finding of Taskonomy by Zamir et al. \cite{ZSSGM18}, by analyzing the benefit of MTL for reducing labeled data to achieve comparable performance to STL, .
%We show that covariate shift, measured by $\Sigma_1^{1/2}\Sigma_2^{-1/2}$, is another cause for suboptimal performance for $\hat{\beta}_t^{\MTL}$.
%		We show that as $n_1 / n_2$ becomes large, having no covariate shift between the source and target tasks yields the optimal performance for the target task.
%		On the other hand, when $n_1 / n_2$ is small, there are counter examples where having the same covariance matrix is not necessarily the optimal choice.

Finally, we connect our theory to practical problems of interest.
We provide a single-task based metric to predict positive or negative transfer in multi-task learning.
While it is not well understood when multi-task learning provides positive transfer, we show that the STL results can help indicate and understand MTL results on ChestX-ray14 \cite{chexnet17} and sentiment analysis datasets \cite{LZWDA18}.
Inspired by our observation that adding more source task data does not always improve performance, we propose to train multi-task models by incrementally adding source task data for predicting a target task.
We show that our training schedule reduces the computational cost by $55\%$ compared to baseline multi-task training on the sentiment analysis dataset, while keeping the accuracy the same.
Last, we validate our three theoretical findings on sentiment analysis tasks.
